---
title: "Boston Marathon: Visualizations"
author: "Me"
date: '2017-10-24'
output: html_document
---

# loading libraries
```{r}
library(tidyverse)
library(grid)
library(RColorBrewer)
library(directlabels)
library(gridExtra)
```


# loading the combined dataset
```{r}
df <- read.csv("~/Desktop/Year 3 Sem 1/LS 88-002/CSVs/combined.csv")
df$official <- as.numeric(as.character(df$official))
df1 <- df[!is.na(df$official), ]
```


# number of runners by year
```{r}
runners_by_year <- df %>%
  filter(year != 2013) %>%
  select(year) %>%
  group_by(year) %>%
  summarise(number_of_runners = n()) %>%
  mutate(diff_2001 = number_of_runners - number_of_runners[1])

runners_by_year_plot <- runners_by_year %>%
  ggplot(aes(x = year, y = number_of_runners, fill = "purple")) + 
  geom_line(aes(group=1), stat = "identity", col = "purple", size = 1.1) +
  labs(title = 'Number of Runners by Year', y = 'Number of Runners', x = "Year") +
  theme(legend.position = "none") + 
  expand_limits(y = 0) +
  scale_x_continuous(breaks = seq(2002, 2014, 2))

runners_by_year_plot
```



# number of runners based on gender
```{r}
gender <- df %>%
  select(year, gender) %>%
  group_by(gender, year) %>%
  summarise(number_of_runners = n())

gender_m <- df %>%
  filter(gender == "M") %>%
  select(year, gender) %>%
  group_by(gender, year) %>%
  summarise(number_of_runners = n())

gender_f <- df %>%
  filter(gender == "F") %>%
  select(year, gender) %>%
  group_by(gender, year) %>%
  summarise(number_of_runners = n()) 

gender_f <- gender_f %>%
  mutate(percentMen = gender_m$number_of_runners / (gender_m$number_of_runners + gender_f$number_of_runners))

gender
gender_m
gender_f
```


# gender plot
```{r}
gender_plot <- gender %>%
  filter(year != 2013) %>%
  ggplot(aes(x = year, y = number_of_runners, group = gender, col = gender)) + 
  geom_line(size = 0.7) + 
  labs(title = 'Number of Runners by Gender', x = "Year", y = 'Number of Runners') +
  geom_dl(aes(label = gender), method = list(dl.combine("last.points"), cex = 1), col = "black") +
  theme(legend.position = "none") + 
  expand_limits(y = 0) +
  scale_x_continuous(breaks = seq(2002, 2014, 2))

gender_plot
```

# gender proportion plot
```{r}
gender_diff_plot <- gender_f %>%
  filter(year != 2013) %>%
  ggplot(aes(x = year, y = percentMen)) + 
  geom_line(col = 'purple', size = 0.9) + 
  labs(title = 'Proportion of Runners Who are Male', x = "Year", y = 'Proportion Male') +
  theme(legend.position = "none") +
  scale_x_continuous(breaks = seq(2002, 2014, 2))

gender_diff_plot
```


# number of runners by age --> shows that runners are getting older over time
```{r}
age_plot1_5 <- df %>%
  filter(year == 2001 | year == 2002 | year == 2003 | year == 2004 | year == 2005) %>%
  filter(age < 85) %>%
  ggplot(aes(x = age)) +
  geom_histogram(binwidth = 10, fill = 'dark red') +
  labs(title = "Ages of Runners from 2001-2005", x = "Age", y = "Count") +
  scale_x_continuous(breaks = c(20, 30, 40, 50, 60, 70, 80))

age_plot6_9 <- df %>%
  filter(year == 2006 | year == 2007 | year == 2008 | year == 2009) %>%
  ggplot(aes(x = age)) +
  geom_histogram(binwidth = 10, fill = 'dark orange') +
  labs(title = "Ages of Runners from 2006-2009", x = "Age", y = "Count") +
  scale_x_continuous(breaks = c(20, 30, 40, 50, 60, 70, 80))

age_plot10_14 <- df %>%
  filter(year == 2010 | year == 2011 | year == 2012 | year == 2013 | year == 2014) %>%
  ggplot(aes(x = age)) +
  geom_histogram(binwidth = 10, fill = 'dark green') +
  labs(title = "Ages of Runners from 2010-2014", x = "Age", y = "Count") +
  scale_x_continuous(breaks = c(20, 30, 40, 50, 60, 70, 80))

age_plot1_5
age_plot6_9
age_plot10_14
```


# age historgrams
```{r}
grid.arrange(age_plot1_5, age_plot6_9, age_plot10_14)
```




# average age
```{r}
age <- df %>%
  select(year, age) %>%
  group_by(year) %>%
  summarise(avg_age = mean(age))
age
```

# holistic age plot
```{r}
age_plot <- age %>% 
  ggplot(aes(x = year, y = avg_age, fill = 'red')) +
  geom_line(aes(group=1), stat = "identity", col = 'orange', size = 0.9) +
  labs(title = 'Average Age by Year', y = "Average Age", x = "Year") +
  scale_x_continuous(breaks = seq(2002, 2014, 2))
age_plot
```




# runtimes by gender
```{r}
runtime <- df1 %>%
  select(year, official) %>%
  filter(year != 2013) %>%
  group_by(year) %>%
  summarise(avg_runtime = mean(official))

runtime_m <- df1 %>%
  filter(gender == "M") %>%
  select(year, official) %>%
  group_by(year) %>%
  summarise(avg_runtime = mean(official))

runtime_f <- df1 %>%
  filter(gender == "F") %>%
  select(year, official) %>%
  group_by(year) %>%
  summarise(avg_runtime = mean(official))

runtime_f <- runtime_f %>%
  mutate(difference = runtime_f$avg_runtime - runtime_m$avg_runtime)

runtime
runtime_m
runtime_f
```


# gender runtime plots
```{r}
runtime_plot <- runtime %>% 
  ggplot(aes(x = year, y = avg_runtime, fill = 'red')) +
  geom_line(aes(group=1), stat = "identity", col = 'purple', size = 0.9) +
  labs(title = 'Average Runtime by Year', y = "Avg Runtime (Minutes)", x = "Year") +  
  geom_smooth(method=lm, se=FALSE, fullrange=FALSE, level=0.95) +
  theme(legend.position = "None") +
  scale_x_continuous(breaks = seq(2002, 2014, 2))
runtime_plot
```

# male runtimes
```{r}
runtime_plot_m <- runtime_m %>% 
  ggplot(aes(x = year, y = avg_runtime, fill = 'red')) +
  geom_line(aes(group=1), stat = "identity", col = 'blue', size = 0.9) +
  labs(title = 'Average Runtime by Year (Male)', y = "Avg Runtime (Minutes)", x = "Year") +
  scale_x_continuous(breaks = seq(2002, 2014, 2)) +
  scale_y_continuous(limits = c(200, 280))
runtime_plot_m
```

#female runtimes
```{r}
runtime_plot_f <- runtime_f %>% 
  ggplot(aes(x = year, y = avg_runtime, fill = 'red')) +
  geom_line(aes(group=1), stat = "identity", col = 'pink', size = 0.9) +
  labs(title = 'Average Runtime by Year (Female)', y = "Avg Runtime (Minutes)", x = "Year") +
  scale_x_continuous(breaks = seq(2002, 2014, 2)) +
  scale_y_continuous(limits = c(200, 280))
runtime_plot_f
```

# male and female runtimes
```{r}
grid.arrange(runtime_plot_m, runtime_plot_f)
```

# gender runtime differences
```{r}
gender_diff <- runtime_f %>%
  ggplot(aes(x = year, y = difference, fill = 'red')) +
  geom_line(aes(group=1), stat = "identity", col = 'purple', size = 0.9) +
  labs(title = 'Average Runtime Difference (Female Minus Male)', y = "Avg Difference (Minutes)", x = "Year") + 
  expand_limits(y = 0) +
  scale_x_continuous(breaks = seq(2002, 2014, 2))
gender_diff
```






# countries
```{r}
countries <- df %>%
  select(year, country) %>%
  group_by(year) %>%
  summarise(num_countries = n_distinct(country))
countries
```


# country plot
```{r}
country_plot <- countries %>%
  ggplot(aes(x = year, y = num_countries, fill = 'red')) +
  geom_line(aes(group=1), stat = "identity", col = 'violet', size = 1) +
  labs(title = 'Number of Countries by Year', y = "Number of Countries", x = "Year") +
  scale_x_continuous(breaks = seq(2002, 2014, 2))  
country_plot
```




# scatterplot of Runtime vs Age
```{r}
age_scatter <- df1 %>%
  ggplot(aes(x = age, y = official, fill = 'red')) +
  geom_point(aes(group=1), stat = "identity", col = 'dark turquoise', size = 1) +
  geom_smooth(method=lm, se=TRUE, fullrange=FALSE, level=0.95) +
  labs(title = 'Runtime Versus Age', y = "Average Runtime", x = "Average Age") +
  theme(legend.position = "none") + 
  annotate(x=20.5, y=1.5, label=paste("R = ", round(cor(df1$age, df1$official),2)), geom="text", size=4.5)
age_scatter
```









